### 2024-08-03 getting-started.qmd
20240802 - remove packages that are no longer needed for this chapter
20240802 - transform between long and wide form

For instance, you can use the following code to install the `nflreadr` package:

@@

install.packages("nflreadr")
```

@@

```{r}
library("ffanalytics")
library("nflreadr")
library("nflfastR")
library("nfl4th")
library("nflplotR")
library("progressr")
library("lubridate")
library("tidyverse")
```

@@

fantasyPoints <- data.frame(
  ID = c(2, 7, 13, 14),
  fantasyPoints = c(250, 170, 65, 15)
)
```

@@

Here are the data in the `nfl_actualStats_offense_weekly` [UPDATE] object.
The data are structured in "player-season-week form".
That is, every row in the dataset is uniquely identified by the variables, `player_id`, `season`, and `week`.
This is an example of long form, because each player has multiple rows.

Original data:

```{r}
# UPDATE

#dataLong <- nfl_actualStats_offense_weekly %>% 
#  select(player_id, player_display_name, season, week, fantasy_points)
#
#dim(dataLong)
#
#names(dataLong)
```

Below, we widen the data by two variables (`season` and `week`), using `tidyverse`, so that the data are now in "player form" (where each row is uniquely identified by the `player_id` variable):

```{r}
# UPDATE

#dataWide <- dataLong %>% 
#  pivot_wider(
#    names_from = c(season, week),
#    names_glue = "{.value}_{season}_week{week}",
#    values_from = fantasy_points)
#
#dim(dataWide)
#
#names(dataWide)
```

@@

Conversely, we can also restructure data from wide to long.

Original data:

```{r}
# UPDATE

#dataWide <- nfl_actualStats_offense_weekly %>% 
#  select(player_id, player_display_name, season, week, recent_team, opponent_team)
#
#dim(dataWide)
#
#names(dataWide)
```

Data in long form, transformed from wide form using `tidyverse`:

```{r}
# UPDATE

#dataLong <- dataWide %>% 
#  pivot_longer(
#    cols = c(recent_team, opponent_team),
#    names_to = "role",
#    values_to = "team")
#
#dim(dataLong)
#
#names(dataLong)
```

>>>>

For instance, you can use the following code to install the `tidyverse` package:

@@

install.packages("tidyverse")
```

@@

```{r}
library("tidyverse")
```

@@

fantasyPoints <- data.frame(
  ID = c(2, 7, 13, 14),
  fantasyPoints = c(250, 170, 65, 15)
)

fantasyPoints_weekly = expand.grid(
  ID = 1:12,
  season = c(2022, 2023),
  week = 1:17
)

set.seed(52242)
fantasyPoints_weekly$fantasyPoints <- sample(
  0:35,
  size = nrow(fantasyPoints_weekly),
  replace = TRUE
)
```

@@

Here are the original data in long form.
The data are structured in "player-season-week form".
That is, every row in the dataset is uniquely identified by the combination of variables, `ID`, `season`, and `week`.
This is an example of long form, because each player has multiple rows.

```{r}
dataLong <- full_join(
  players %>% select(-age),
  fantasyPoints_weekly,
  by = c("ID")
)

dataLong

dim(dataLong)

names(dataLong)
```

Below, we widen the data by two variables (`season` and `week`), using `tidyverse`, so that the data are now in "player form" (where each row is uniquely identified by the `ID` variable):

```{r}
dataWide <- dataLong %>% 
  pivot_wider(
    names_from = c(season, week),
    names_glue = "{.value}_{season}_week{week}",
    values_from = fantasyPoints)

dataWide

dim(dataWide)

names(dataWide)
```

@@

Conversely, we can also restructure data from wide to long.
Here are the data in long form, after they have been transformed from wide form using `tidyverse`:

```{r}
dataLong <- dataWide %>% 
  pivot_longer(
    cols = fantasyPoints_2022_week1:fantasyPoints_2023_week17,
    names_to = c("season", "week"),
    names_pattern = "fantasyPoints_(.*)_week(.*)",
    values_to = "fantasyPoints")

dataLong

dim(dataLong)

names(dataLong)
```
